Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-02-18 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: added todo and spell-checked

% HS: temporary title. Feel free to add suggestions
% \title{Probing higher-curvature gravity theories from binary BH ringdown signals}
\title{Black-hole ringdown as a probe of higher-curvature gravity theories}

@@

The formalism modifies the Kerr quasinormal mode frequency as
%
\begin{subequations}
\begin{align}
\omega_{\iota} &= \omega^{\rm K}_{\iota} \, (1 + \delta \omega_{\iota}), \\
\tau_{\iota}   &= \tau^{\rm K}_{\iota}   \, (1 + \delta \tau_{\iota}),
\end{align}
\label{eq:general_deviation}
\end{subequations}
%
where the $\iota$ collectively represents the QNM labels $\{l,\, m,\, n\}$.

@@

% ---------------------------
% CONCRETE EXAMPLE OF MAPPING
% ---------------------------
Let us show one concrete example of how this mapping is done.
%
In Ref.~\cite{Wagle:2021tam} the QNMs of slow-rotating BHs in dCS gravity were calculated
and it was found that the axial gravitational modes have their damping time increased as we
increase the Chern-Simons coupling $\ell_{\rm CS}$ at fixed spin $\chi$ of the BH.
%
Having selected this mode we can now obtain $\delta\omega^{(i)}$ and $\delta\tau^{(i)}$ as follows.
%
Using the fitting formula Eq.~(54a) of~\cite{Wagle:2021tam}, namely,
%
\begin{equation}
    M \omega^{\rm CS}_{220} = c_{1} + c_{2} \kappa \zeta + (c_{3} + c_{4} \kappa \zeta) \, (1 - \chi)^{c_{5} + c_{6} \kappa \zeta},
    \label{eq:omega_dsc_eg}
\end{equation}
%
and similarly for the imaginary part ${\rm Im}(\sigma_{220}) = \omega^{\rm CS}_{220} = 1 / \tau^{\rm CS}_{220}$.
%
Here $\kappa = 1/(16 \pi)$, $\zeta = \ell_{\rm CS}^{4} / (M^4 \kappa)$ [thus $\kappa \zeta = (\ell_{\rm CS} / M)^4$]
and $c_{i}$ are fitting coefficients.
%
In the context of a binary BH merger, $M$ is interpreted as the remnant mass $\Mf$.

@@

\begin{table*}[th]
\begin{tabular}{c | c c c c c c c c}
\hline
\hline
Theory & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & QNM & Constraint & This work \\
\hline
sGB      & 4 & 0.0107 & 0.0044 & \dots & \dots &  \cite{Pierini:2021jxd} & $\ell_{\rm GB} \leqslant 1.7$~km~\cite{Perkins:2021mhb} (GW) & \dots \\
dCS      & 4 & 3.1964 & 6.3619 & \dots & \dots &  \cite{Wagle:2021tam}   & $\ell_{\rm CS} \leqslant 8.5$~km~\cite{Silva:2020acr} (EM+GW)  & \dots \\
cEFTofGR & 4 & $-0.3680$ $(-0.5131)$  & 1.9242 (1.7000) & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
qEFTofGR & 6 & \dots & \dots & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
\hline
\hline
\end{tabular}
\caption{Summary of the quasinormal modes calculations.
%
We summarize each theory we have considered together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(i)}_{220}$ and decay time $\delta \tau^{(i)}_{220}$, the
references from which we used the results from and the current best constraint
(if applicable).
%
In the entries for cEFTofGR, the first entries in the $\delta
\omega^{(i)}_{220}$, $\delta \tau^{(i)}_{220}$ are obtained by considering
\emph{only parity-even corrections} and the second entries by considering
\emph{only parity-odd corrections}. In these cases, the dimensionless parameters
$\lambda_{\rm e,o}$ are degenerate with the length scale $\ell$.
%
\hs{Add a column with the constraint (if any) that we placed.}
\hs{TODO: add which runs are done and which need to be done.}
}
\label{tab:ref_theories_qnms}
\end{table*}

@@

In this section, we provide a basic outline of the Bayesian formalism we use to infer the properties of the underlying GW signal, identify the most promising events from the catalog of GW observations to base our analyses on, and discuss the key insights they provide in constraining predictions of momdified gravity theories.

@@

\subsubsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately described by our waveform model \pSEOB{}, we can infer the parameters of the model, $\bm{\lambda}$, given the hypotheis $\mathcal{H}$, using the Bayes theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence. The set of parameters, $\bm{\lambda}$, detailed in Sec.~\ref{sec:review_pSEOB}
include the set of beyond-GR parameters, $\{\ell, \delta \omega_k^{(j)}, \delta \tau_k^{(j)}\}$.
%
As discussed in Secs.~\ref{sec:review_theories} and \ref{sec:theory_specific_qnm}, there is a single coupling
constants in the cases of EdGB and dCS theories, but two in the cubic and
quartic EFTofGR. We also hold the other beyond-GR parameters, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$
fixed to theory-specific predictions. Hence, our set of beyond--GR parameters only includes the parameter $\{\ell \}$.
%

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln \mathcal{L}(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk
denotes complex conjugation and $S_{n}(f)$ is the power spectrum density
of the detector. Assuming a speicifc prior distribution for our parameters, 
we stochastically sample over the parameter space 
using a \texttt{LALInference} Markhov Chain Monte Carlo 
algorithm.~\cite{lallsuite}. We subsequently marginalise over the 
remaining parameters to obtain the posterior probability distribution on $\ell$. 
An example of this is shown in Fig. XX.

@@

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/dcs_GW200129.pdf}
\caption{Posterior probability distribution on $\ell_{\rm CS}$ with GW200129.
The posterior shows that information was gained on this parameter relative to
the initial uniform prior distribution $\ell_{\rm CS} \in [0,\, 200]$~km.
%
The vertical lines represent two threshold for the validity of the theory as
an EFT.
%
The dotted line (labelled ``$m_2$'') at $\ell_{\rm CS} \approx 21.3$~km is
given by the median value of the secondary's mass $\varepsilon \, (G m_2 / c^2)$.
%
The dot-dashed line (labelled ``$M_{f}$'') at $\ell_{\rm CS} \approx 43.8$~km is
given by the median value of the remnant's mass $\varepsilon \, (G M_{f} / c^2)$.
%
For both, we used $\varepsilon = 0.5$.
%
If we take the latter as a length scale for the validity of dCS as an EFT, we
have place upper bound on $\ell_{\rm CS} \leqslant 43.8$~km, within the
approximation of our formalism.
}
\label{fig:dCS_bounds}
\end{figure}

>>>>

% HS: temporary title. Feel free to add suggestions
\title{Black-hole ringdown as a probe of higher-curvature gravity theories}

@@

The formalism modifies the Kerr quasinormal mode frequency as
%
\begin{subequations}
\begin{align}
\omega_{J} &= \omega^{\rm K}_{J} \, (1 + \delta \omega_{J}), \\
\tau_{J}   &= \tau^{\rm K}_{J}   \, (1 + \delta \tau_{J}),
\end{align}
\label{eq:general_deviation}
\end{subequations}
%
where the $J$ collectively represents the QNM labels $\{l,\, m,\, n\}$.

@@

% ---------------------------
% CONCRETE EXAMPLE OF MAPPING
% ---------------------------
Let us show one concrete example of how this mapping is done.
%
In Ref.~\cite{Wagle:2021tam} the QNMs of slow-rotating BHs in dCS gravity were calculated
and it was found that the axial gravitational modes have their damping time increased as we
increase the Chern-Simons coupling $\ell_{\rm CS}$ at fixed spin $\chi$ of the BH.
%
Having selected this branch of QNMs we can proceed to determine $\delta\omega^{(i)}$ and $\delta\tau^{(i)}$ as follows.
%
Using the fitting formula Eq.~(54a) of~\cite{Wagle:2021tam}, namely,
%
\begin{equation}
    M \omega^{\rm CS}_{220} = c_{1} + c_{2} \kappa \zeta + (c_{3} + c_{4} \kappa \zeta) \, (1 - \chi)^{c_{5} + c_{6} \kappa \zeta},
    \label{eq:omega_dsc_eg}
\end{equation}
%
and similarly for the imaginary part ${\rm Im}(\sigma_{220}) = \omega^{\rm CS}_{220} = 1 / \tau^{\rm CS}_{220}$.
%
Here $\kappa = 1/(16 \pi)$, $\zeta = \ell_{\rm CS}^{4} / (M^4 \kappa)$ [thus $\kappa \zeta = (\ell_{\rm CS} / M)^4$]
and $c_{i}$ are fitting coefficients.
%
In the context of a binary BH merger, $M$ is interpreted as the remnant mass $\Mf$.

@@

\begin{table*}[th]
\begin{tabular}{c | c c c c c c c c}
\hline
\hline
Theory & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & QNM & Constraint & This work \\
\hline
sGB      & 4 & 0.0107 & 0.0044 & \dots & \dots &  \cite{Pierini:2021jxd} & $\ell_{\rm GB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & \dots \\
dCS      & 4 & 3.1964 & 6.3619 & \dots & \dots &  \cite{Wagle:2021tam}   & $\ell_{\rm CS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & \dots \\
cEFTofGR & 4 & $-0.3680$ $(-0.5131)$  & 1.9242 (1.7000) & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
qEFTofGR & 6 & \dots & \dots & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
\hline
\hline
\end{tabular}
\caption{Summary of the quasinormal modes calculations.
%
We summarize each theory we have considered together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(i)}_{220}$ and decay time $\delta \tau^{(i)}_{220}$, the
references from which we used the results from and the current best constraint
(if applicable).
%
In the entries for cEFTofGR, the first entries in the $\delta
\omega^{(i)}_{220}$, $\delta \tau^{(i)}_{220}$ are obtained by considering
\emph{only parity-even corrections} and the second entries by considering
\emph{only parity-odd corrections}. In these cases, the dimensionless parameters
$\lambda_{\rm e,o}$ are degenerate with the length scale $\ell$.
%
\hs{Add a column with the constraint (if any) that we placed.}
\hs{TODO: add which runs are done and which need to be done.}
}
\label{tab:ref_theories_qnms}
\end{table*}

@@

In this section, we provide a basic outline of the Bayesian formalism we use to
infer the properties of the underlying GW signal, identify the most promising
events from the catalog of GW observations to base our analyses on, and discuss
the key insights they provide in constraining predictions of modified gravity
theories.

@@

\subsubsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using the Bayes
theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence. The set of parameters, $\bm{\lambda}$, detailed in Sec.~\ref{sec:review_pSEOB}
include the set of beyond-GR parameters, $\{\ell, \delta \omega_k^{(j)}, \delta \tau_k^{(j)}\}$.
%
As discussed in Secs.~\ref{sec:review_theories} and \ref{sec:theory_specific_qnm}, there is a single coupling
constants in the cases of EdGB and dCS theories, but two in the cubic and
quartic EFTofGR. We also hold the other beyond-GR parameters, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$
fixed to theory-specific predictions. Hence, our set of beyond--GR parameters only includes the parameter $\{\ell \}$.
%

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln P(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters, we stochastically
sample over the parameter space using a \texttt{LALInference} Markov Chain
Monte Carlo algorithm.~\cite{lalsuite}. We subsequently marginalize over the
remaining parameters to obtain the posterior probability distribution on
$\ell$. An example of this is shown in Fig. XX.

@@

\hs{Note to ourselves: as mentioned in~\cite{Maselli:2019mjd}:
%
``As customary in parametrized approaches, we focus on perturbative corrections by assuming $\gamma_{i} \delta \omega^{(n)} \ll 1$, $\gamma_{i} \delta \tau^{(n)} \ll 1$ and GR is recovered in the limit $\gamma_{i} \to 0$.''
We should check this \emph{in addition} to the small-coupling approximation to argue whether we placed a bound or not.
}

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/dcs_GW200129.pdf}
\caption{Posterior probability distribution on $\ell_{\rm CS}$ with GW200129.
The posterior shows that information was gained on this parameter relative to
the initial uniform prior distribution $\ell_{\rm CS} \in [0,\, 200]$~km.
%
The vertical lines represent two threshold for the validity of the theory as
an EFT.
%
The dotted line (labeled ``$m_2$'') at $\ell_{\rm CS} \approx 21.3$~km is
given by the median value of the secondary's mass $\varepsilon \, (G m_2 / c^2)$.
%
The dot-dashed line (labeled ``$M_{f}$'') at $\ell_{\rm CS} \approx 43.8$~km is
given by the median value of the remnant's mass $\varepsilon \, (G M_{f} / c^2)$.
%
For both, we used $\varepsilon = 0.5$.
%
If we take the latter as a length scale for the validity of dCS as an EFT, we
have place upper bound on $\ell_{\rm CS} \leqslant 43.8$~km, within the
approximation of our formalism.
}
\label{fig:dCS_bounds}
\end{figure}
